A New Method of Modelling Tuneable Lasers with Functional Composition

### 2019-05-28 Thesis/Nonlinear/Nonlinear.tex
Finished corrections

\subsection{Validation}
Before delving into the nonlinear behaviour of the model, we wish to validate that the code is working as expected---and as a sanity test for the linear solution---by comparing the results of the simulations with $b=0$ to the results of the linear model in Chapter \ref{chap:linear}. In the case of the numerical solution our initial conditions become somewhat important. In all of the following analysis the initial waveform is $\Gamma \sech \left( 2T \right) \textrm{e}^{i \pi / 4}$ normalized so that the initial energy is $E_0 = 0.1$; additionally, $a = 8000$, and $h = 0.04$. \\
\begin{figure}[tbp]
\centering
\input{./Figures/Variance}
\caption{Simulation and analytic equilibrium variance, chirp, and phase shift as a function of $s$.}
\label{fig:var}
\end{figure}

@@

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique Batman-like shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---this will be a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, this is what we find for moderate values of $T$. On the other hand, for $|T| > 1$ the relation begins to level off. This is in agreement with experimental results \cite{chen, rothenberg, tomlinson}. \\

@@

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. The energy of the pulse at equilibrium\footnote{The pulse is passed through each component of the laser 40 times before the energy is computed.} is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$, and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy and discontinuous. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary appears to become periodic as $b$ increases around $s = 0.3$. \\

>>>>

\subsection{Validation}
Before delving into the nonlinear behaviour of the model, we wish to validate that the code is working as expected---and as a sanity test for the linear solution---by comparing the results of the simulations with $b=0$ to the results of the linear model in Chapter \ref{chap:linear}. In the case of the numerical solution our initial conditions become somewhat important. In all of the following analysis the initial waveform is $\Gamma \sech \left( 2T \right) \textrm{e}^{i \pi / 4}$ normalized so that the initial energy is $E_0 = 0.1$; additionally, $a = 8000$, and $h = 0.04$. \\
\begin{figure}[tbp]
\centering
\input{./Figures/Variance}
\caption{Simulation and analytic equilibrium variance, chirp, and phase shift as a function of $s$. The analytic expressions are given by \eqref{eq:equilvar}, \eqref{eq:equilchirp}, \eqref{eq:phase}, respectively.}
\label{fig:var}
\end{figure}

@@

\section{Nonlinear Model}
\begin{figure}[tbp]
\centering
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Shape}
\caption{Envelope}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_FT}
\caption{Fourier transform}
\end{subfigure} \\
\begin{subfigure}{0.5\textwidth}
\centering
\input{./Figures/Stable_Chirp}
\caption{Chirp}
\end{subfigure}
\caption[Envelope, Fourier transform, and chirp of the pulse---stable case.]{Simulation with $s = 0.15$, and $b = 2.1$ after 15 circuits.}
\label{fig:nlstable}
\end{figure}
With the inclusion of the nonlinearity we generally find a similar solution to the linear case. An example of this is shown in Figure \ref{fig:nlstable}. The envelope of the pulse is unsurprisingly Gaussian-esque, however, it is \emph{not} precisely Gaussian and more closely resembles a generalized Gaussian\footnote{A generalized Gaussian has the form $\exp \left(-t^\alpha \right)$, with $\alpha > 2$.}. The fact that the pulse envelope is not Gaussian is further emphasized while examining the Fourier transform of the pulse. If the pulse were a Gaussian, we would expect the Fourier transform to also be a Gaussian \cite{debnath, gradshteyn}. Instead, the magnitude of the Fourier transform has a unique Batman-like shape. This deviation suggests the nonlinearity implants higher frequency oscillations into the pulse---this will be a key observation in the coming subsections. Finally, we shall examine the derivative of the phase---essentially the chirp. Recall in Chapter \ref{chap:linear} the chirp was defined as coefficient of $-\frac{1}{2}iT^2$ in the exponential, by taking the negative of the derivative we would expect a linear function with a slope equal to the chirp. In the nonlinear case, this is what we find for moderate values of $T$. On the other hand, for $|T| > 1$ the relation begins to level off. This is in agreement with experimental results \cite{chen, rothenberg, tomlinson}. While the nonlinear model yields similar results as the linear model (Figure \ref{fig:linear}), very interesting variations arise from the nonlinearity. \\

@@

\begin{figure}[tbp]
\input{./Figures/Cartoon}
\caption{Stability of the pulse in the $s$-$b$ plane.}
\label{fig:cartoon}
\end{figure}

The energy of the pulse is directly related to the output power of the laser, and since this is not as controllable as in a regular laser, it is of great interest. The energy of the pulse at equilibrium\footnote{The pulse is passed through each component of the laser 40 times before the energy is computed.} is shown in Figure \ref{fig:energy}. In Figure \ref{fig:energyzoom} we generally find that the energy is a smooth surface that slowly decays as $s$, and $b$ increase. In the upper left hand region of the plot this is clearly not the case. The contours show that the energy is very noisy and discontinuous. Perhaps more surprising, as shown in Figure \ref{fig:energybig}, is that this boundary appears to become periodic as $b$ increases around $s = 0.3$. These features are further highlighted in Figure \ref{fig:cartoon} by disregarding the energy and only examining whether the pulse is stable or not. \\

### 2019-05-28 Thesis/Previous_Models/Previous.tex
Finished corrections

\chapter{Previous Modelling Efforts}
In this chapter we shall explore the previous modelling efforts by first reviewing the classic equations used in nonlinear optics. We will then build upon this to the master equation of mode-locking, where the solutions to this will briefly be studied before diving into `discretized' functional models. \\

\section{Generalized Nonlinear Schr\"odinger Equation}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE) \cite{agrawal2013, anderson, burgoyne2007, desurvire, ferreira, finot, rothenberg}
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
\end{align*}
Here $A$ is the complex pulse amplitude, $\beta_2$ is the second-order---or group delay---dispersion, and $\gamma$ is the coefficient of nonlinearity or Kerr coefficient. In practice, this equation lacks a few key terms. Thus, it is often generalized by adding amplification, loss, and higher order terms. This gives the generalized nonlinear Schr\"odinger equation (GNLSE) \cite{agrawal2013, bohun, finot, peng, shtyrina, yarutkina},
\begin{align}
\label{eq:nlse}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \frac{\beta_3}{6}\pdiff[3]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
\end{align}
where $\beta_3$ is the third order coefficient of dispersion, $g(A)$ is an amplifying term due to the gain, and $\alpha$ is the loss due to scattering and absorption. \\

This equation can be derived from the nonlinear wave equation, the derivation is presented in detail in \cite{agrawal2013, ferreira}. Within the derivation comoving coordinates are used so that the reference frame propagates with the pulse at the group velocity. This is achieved with the substitution
\begin{align*}
T = t - \frac{z}{v_g}.
\end{align*}

The GNLSE takes the same form as the Schr\"odinger equation with the inclusion of the cubic nonlinear term, hence its name. For this reason, it is sometimes referred to as the cubic nonlinear Schr\"odinger equation. For intensities approaching $1 \text{ GW/cm}^2$, the $\gamma$ parameter must be replaced by $\gamma_0 (1 - b_s |A|^2)$, where $b_s$ is a saturation parameter \cite{agrawal2013}, this has the addition of a quintic term to incorporate nonlinearities associated with such large powers. Furthermore, the $\beta$ terms come from a Taylor expansion of the wavenumber \cite{kartner}, that is,
\begin{align*}
k(\omega) &= k_0 + \pdiff{k}{\omega}(\omega - \omega_0) + \frac{1}{2} \pdiff[2]{k}{\omega}(\omega - \omega_0)^2 + \frac{1}{6} \pdiff[3]{k}{\omega}(\omega - \omega_0)^3 + \dots, \\
&= \phi + \frac{1}{v_g}(\omega - \omega_0) + \frac{1}{2}\beta_2 (\omega - \omega_0)^2 + \frac{1}{6}\beta_3 (\omega - \omega_0)^3 + \dots,
\end{align*}
where $\phi$ is the phase shift. Typically, the third order effects must only be considered for ultrashort pulses---pulse widths less than $\sim5 \text{ ps}$---because of their large bandwidth \cite{agrawal2013}.

\section{The Master Equation of Mode-Locking}
The GNLSE has many applications in nonlinear optics and fibre optic communications, however, in the context of lasers we wish to add a modulation term to ensure mode-locking, this yields the master equation of mode-locking, \cite{hausbook, haus1975, haus1986, haus1992, haus2000, kartner, tamura, usechak}
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \frac{\beta_3}{6}\pdiff[3]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - M(T).
\end{align*}
The most common form of modulation is the sinusoid $M(T) = \frac{1}{2}\frac{M_s}{\omega_M} \left( 1 - \cos \left( \omega_M T \right) \right)$ \cite{hausbook, haus1975, haus1996, kartner}, and since $T$ is generally small, by it's definition, we expand via it's Taylor series so that $M(T) = \frac{1}{2}M_s T^2$. This brings us to the most common form of the master equation of mode-locking (neglecting the third order dispersion),
\begin{align}
\label{eq:meml}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - \frac{1}{2}M_s T^2.
\end{align}
Commonly, in an attempt to simplify the equation the gain assumed to be constant. As a whole no analytic solution is know for \eqref{eq:meml}, however, after additional simplifications there are three flavours of solutions. \\

In the least complicated case, the modulation and nonlinearity are omitted resulting in a solution in the form of a hyperbolic secant \cite{haus1975, haus1986, haus1992}. With the inclusion of the nonlinearity a similar solution is found, however, it is now it the form of a chirped hyperbolic secant \cite{haus1991, usechak}. Both of these are unsurprising since this reduces to the soliton solution of the NLSE \cite{ferreira}. Lastly, by including the modulation term, and excluding the nonlinearity, \eqref{eq:meml} can be solved using separation of variables and one finds the solutions are the Gaussian--Hermite polynomials \cite{burgoyne2014, hausbook, haus1975, haus1996, haus2000, kartner, tamura, usechak}. However, in practice only the Gaussian is stable---the higher modes quickly decay within the laser \cite{hausbook, haus1975, haus1996, haus2000}. For a more comprehensive and exhaustive history see \cite{haus2000}. \\

@@

Such a method was first proposed in 1955 by Cutler \cite{cutler} while analyzing a microwave regenerative pulse generator. This method was adapted for mode-locked lasers in 1969 by Siegman and Kuizenga \cite{seigman}, which they then greatly added to the following year \cite{kuizenga1970a}. Kuizenga and Seigman also had success experimentally validating their model \cite{kuizenga1970b, kuizenga1970}. The effects of the nonlinearity would not be considered until Martinez, Fork, and Gordon \cite{martinez1984, martinez1985} tried to model passive mode-locking---mode-locking without the use of a modulator. In the absence of a modulator the nonlinearity becomes crucial to shaping the pulse. I don't know how to start this sentence*** when it was resurrected by Burgoyne \cite{burgoyne2014} in 2014 for tuneable lasers. \\

Despite these attempts, several short-comings exist. The clearest is that none of these models have contained every block--either the nonlinearity or the modulation have been omitted. In the framework of tuneable lasers each component plays an important role. Another key drawback is that the functional operations of some of the components are somewhat phenomenological. While these functions are chosen based on the observed output, they are not necessarily correct. Finally, none of these previous models have been able to exhibit a phenomena called \emph{wave-breaking} in which the self-phase modulation (SPM) of the pulse becomes too strong distorting and damaging the wave until it ultimately becomes unstable and unsustainable. This notion will be explained in greater detail in Chapter \ref{chap:nl}. \\

>>>>

\chapter{Previous Modelling Efforts}
In this chapter we shall explore the previous modelling efforts by first reviewing the classic equations used in nonlinear optics. We will then build upon this to obtain the master equation of mode-locking, where the solutions to this will briefly be reviewed before considering `discretized' functional models. \\

\section{Generalized Nonlinear Schr\"odinger Equation}
The standard equation for studying nonlinear optics is the nonlinear Schr\"odinger equation (NLSE) \cite{agrawal2013, anderson, burgoyne2007, desurvire, ferreira, finot, rothenberg}
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A.
\end{align*}
Here $A(z, T)$ is the complex pulse amplitude, $\beta_2 \in \mathbb{R}$ is the second-order---or group delay---dispersion, and $\gamma \in \mathbb{R}$ is the coefficient of nonlinearity or Kerr coefficient. This equation can be derived from the nonlinear wave equation for electric fields, the derivation is presented in detail in \cite{agrawal2013, ferreira}. Within the derivation comoving coordinates are used so that the reference frame propagates with the pulse at the group velocity. This is achieved with the substitution
\begin{align}
\label{eq:shift}
T = t - \frac{z}{v_g}.
\end{align}
In this way, $z$ defines the distance travelled by the pulse, and $T$ the time difference from the peak of the pulse. Moreover, $T = 0$ is the peak of the pulse regardless of the actual time elapsed, because of this, generally $T$ is not very large. In standard coordinates the nonlinear Schr\"odinger equation has an additional term
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{t} + i \gamma |A|^2 A - \frac{1}{v_g} \pdiff{A}{t}.
\end{align*}
However, under the transformation \eqref{eq:shift}, we have that $\pdiff{}{t} \mapsto \pdiff{}{T}$, and $\pdiff{}{z} \mapsto \pdiff{}{z} - \frac{1}{v_g} \pdiff{}{T}$ by chain rule, thus, cancelling out this final term. \\

In practice, this equation lacks a few key terms. Thus, it is often generalized by adding amplification, loss, and higher order terms. This gives the generalized nonlinear Schr\"odinger equation (GNLSE) \cite{agrawal2013, bohun, finot, peng, shtyrina, yarutkina},
\begin{align}
\label{eq:nlse}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \frac{\beta_3}{6}\pdiff[3]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A,
\end{align}
where $\beta_3 \in \mathbb{R}$ is the third order coefficient of dispersion, $g(A)$ is an amplifying term due to the gain, and $\alpha \in \mathbb{R}$ is the loss due to scattering and absorption. \\

The GNLSE takes the same form as the Schr\"odinger equation with the inclusion of the cubic nonlinear term, hence its name. For this reason, it is sometimes referred to as the cubic nonlinear Schr\"odinger equation. For intensities approaching $1 \text{ GW/cm}^2$, the $\gamma$ parameter must be replaced by $\gamma_0 (1 - b_s |A|^2)$, where $b_s$ is a saturation parameter \cite{agrawal2013}, this has the addition of a quintic term to incorporate nonlinearities associated with such large powers. Furthermore, the $\beta$ terms come from a Taylor expansion of the wavenumber \cite{kartner}, that is,
\begin{align*}
k(\omega) &= k_0 + \left. \pdiff{k}{\omega}\right|_{\omega_0}(\omega - \omega_0) + \frac{1}{2} \left. \pdiff[2]{k}{\omega}\right|_{\omega_0}(\omega - \omega_0)^2 + \frac{1}{6} \left. \pdiff[3]{k}{\omega}\right|_{\omega_0}(\omega - \omega_0)^3 + \cdots \\
&= k_0 + \frac{1}{v_g}(\omega - \omega_0) + \frac{1}{2}\beta_2 (\omega - \omega_0)^2 + \frac{1}{6}\beta_3 (\omega - \omega_0)^3 + \cdots,
\end{align*}
where $k_0$ is the phase shift. Typically, the third order effects must only be considered for ultrashort pulses---pulse widths less than $\sim5 \text{ ps}$---because of their large bandwidth~\cite{agrawal2013}.

\section{The Master Equation of Mode-Locking}
\label{chap:meml}
The GNLSE has many applications in nonlinear optics and fibre optic communications, however, in the context of lasers we wish to add a modulation term to ensure mode-locking, this yields the master equation of mode-locking, \cite{hausbook, haus1975, haus1986, haus1992, haus2000, kartner, tamura, usechak}
\begin{align}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \frac{\beta_3}{6}\pdiff[3]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - M(T).
\end{align}
The most common form of modulation is the sinusoid $M(T) = \frac{1}{2}\frac{M_s}{\omega_M^2} \left( 1 - \cos \left( \omega_M T \right) \right)$ \cite{hausbook, haus1975, haus1996, kartner}, and since $T$ is generally small, by \eqref{eq:shift}, we expand via it's Taylor series so that $M(T) = \frac{1}{2}M_s T^2 + \bigO{T^4}$. This brings us to the most common form of the master equation of mode-locking (neglecting the third order dispersion),
\begin{align}
\label{eq:meml}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \frac{1}{2}g(A) A - \alpha A - \frac{1}{2}M_s T^2.
\end{align}
Commonly, in an attempt to simplify, the equation the gain assumed to be constant with $g(A) = 2g_0 \in \mathbb{R}$. As a whole no analytic solution is know for \eqref{eq:meml}, however, after additional simplifications there are three flavours of solutions. \\

In the least complicated case, the modulation and nonlinearity are omitted giving
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \left( g_0 - \alpha \right) A.
\end{align*}
This results in a solution in the form of a hyperbolic secant \cite{haus1975, haus1986, haus1992}. On the other hand, including the nonlinearity yields
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + i \gamma |A|^2 A + \left( g_0 - \alpha \right) A,
\end{align*}
where a similar solution is found, however, it is instead of the form of a chirped hyperbolic secant \cite{haus1991, usechak}. Both of these are unsurprising since this reduces to the soliton solution of the NLSE \cite{ferreira}. Finally, by including the modulation term, and excluding the nonlinearity we have
\begin{align*}
\pdiff{A}{z} &= - i \frac{\beta_2}{2}\pdiff[2]{A}{T} + \left( g_0 - \alpha \right) A - \frac{1}{2}M_s T^2.
\end{align*}
This can be solved using separation of variables and one finds the solutions are the Gaussian--Hermite polynomials \cite{burgoyne2014, hausbook, haus1975, haus1996, haus2000, kartner, tamura, usechak}. However, in practice only the Gaussian is stable---the higher modes quickly decay within the laser \cite{hausbook, haus1975, haus1996, haus2000}. For a more comprehensive and exhaustive history see \cite{haus2000}. \\

@@

Such a method was first proposed in 1955 by Cutler \cite{cutler} while analyzing a microwave regenerative pulse generator. This method was adapted for mode-locked lasers in 1969 by Siegman and Kuizenga \cite{seigman}, which they then greatly added to the following year \cite{kuizenga1970a}. Kuizenga and Seigman also had success experimentally validating their model \cite{kuizenga1970b, kuizenga1970}. The effects of the nonlinearity would not be considered until Martinez, Fork, and Gordon \cite{martinez1984, martinez1985} tried to model passive mode-locking---mode-locking without the use of a modulator. In the absence of a modulator the nonlinearity becomes crucial to shaping the pulse. This issue has recently been readdressed by Burgoyne \cite{burgoyne2014} in the literature for tuneable lasers. In each of these models the effect of each block is described by a Gaussian transfer function. These transfer functions are then multiplied together to give the overall effect. \\

Despite these attempts, several short-comings exist. The clearest is that none of these models have contained every block--either the nonlinearity or the modulation have been omitted. In the framework of tuneable lasers, each component plays a crucial role and the tuneable laser will not function without the inclusion of all of the components. Another key drawback is that the functional operations of some of the components are somewhat phenomenological. While these functions are chosen based on the observed output, they are not necessarily consistent with their underlying physics. Finally, none of these previous models have been able to exhibit a phenomena called \emph{wave-breaking} in which the self-phase modulation (SPM) of the pulse becomes too strong, distorting and damaging the wave until it ultimately becomes unstable and unsustainable. This notion will be explained in greater detail in Chapter \ref{chap:nl}. \\
